---
title: "Lab 3a - Exercises"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Fall 2025"
output: pdf_document
---

## Set up the session
- Load packaged: tidyverse, lubridate, & here
- Import the `EPAair_PM25_NC2019_raw.csv` file
  - Set strings to be read in as factors
  - Set the "Site.ID" column to be read in as a factor
- Set the date column to be a date

```{r Setup}
#Import packages
library(tidyverse); library(lubridate); library(here)

#Check workspace
here()

#Import data
pm25 <- read.csv(
  file = here ("./Data/Raw/EPAair_PM25_NC2019_raw.csv"),
  stringsAsFactors = TRUE,
  colClasses = c(Site.ID = "factor")
)

#Fix the date
pm25$Date <- mdy(pm25$Date)

```


## Explore properties of the dataframe
-   How many different sites are included in the data set?
-   What is the mean daily PM2.5 concentration?
-   In what units of are PM 2.5 concentrations reported?

```{r Explore the data}
# Compute the number of different sites
length(unique(pm25$Site.Name))
pm25 %>% summarize(n_sites = n_distinct(Site.Name))


# Compute the mean and median PM 2.5 concentration
mean(pm25$Daily.Mean.PM2.5.Concentration, na.rm = TRUE)

median(pm25$Daily.Mean.PM2.5.Concentration, na.rm = TRUE)

# What are the units

unique(pm25$Units)

# What is the date range (min and max dates)
min(pm25$Date, na.rm = TRUE)

max(pm25$Date, na.rm = TRUE)
```


## Plot distribution of values

-   Number of samples using each different source 
-   Distribution of daily mean concentration values
-   Distribution of daily mean concentration, by source*
  *   There are multiple ways to visualize this...

```{r Visualize the data}
#Number of samples using each different source 
table(pm25$Site.Name)

pm25$Site.Name <- factor(pm25$Site.Name)
summary(pm25$Site.Name)

#Distribution of daily mean concentration values
hist(pm25$Daily.Mean.PM2.5.Concentration,
     main = "Distribution of Daily Mean PM2.5 Concentration",
     xlab = "PM2.5 Concentration")

#Distribution of daily mean concentration, by source

# Make sure Site.Name is a factor 
pm25$Site.Name <- factor(pm25$Site.Name)

boxplot(Daily.Mean.PM2.5.Concentration ~ Site.Name,
        data = pm25,
        main = "Distribution of Daily Mean PM2.5 by Site",
        xlab = "Monitoring Site",
        ylab = "PM2.5 Concentration")

```

